### 2018-03-19 submission/manuscript.Rmd
discussion and conclusion updates

These observations build upon previous studies [@taq_Gohl_2016] by showing that different HiFi DNA polymerases have significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5]. HiFi DNA polymerase dependent differences in total number of OTUs are important to consider since many diversity metrics rely on this measure as part of their calculations. Our observations show that HiFi DNA polymerase can have a noticeable affect on the OTUs generated and these differences are consistent across sub-sampling depth and PCR cycle number [Figure 2-4]. Importantly, high biomass samples may underestimate the biases present within low biomass samples. An indication that this may be the case is the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples.

Although we did not observe strong differences in the Bray-Curtis index the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30 cycles, that are commonly used. There was few differences within individuals between corresponding 5-cycle increments (e.g. 15 to 20, 20 to 25, etc.). However, there is a clear trend that suggests that the community at 20 cycles is different then the community at 25 cycles [Figure 3]. These findings, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached. Further, for some HiFI DNA polymerases, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5]. The chimera prevalence was strongly correlated with the number of OTUs which is relied upon heavily for many community metric calculations. However, Bray-Curtis analysis with PERMANOVA showed few differences based on HiFi DNA polymerase. It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully down-weight these respective OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases.

Our observations suggest that there are clear HiFi DNA polymerase dependent differences in both per base error rate and chimera prevalence that cannot be removed using bioinformatic approaches [Figure 4 & 5]. Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A family polymerase [@polymerase_Ishino_2014]. In fact, from the material safety data sheet (MSDS), the differences between the two is not immediately apparent. Both Accuprime and Platinum contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. It is possible that differences in how the recombinant *Taq* was generated could be the main reason for the differences in chimera rate.

@@

Our findings show that the number of OTUs will be dependent on both HiFi DNA polymerase and cycle number chosen. Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.

>>>>

In this study we show that the number of OTUs, error rate, and chimera prevalence are HiFi DNA polymerase dependent [Figure 1-2 & 4-5]. These differences are important because many diversity metrics rely on this measure as part of their calculations (e.g. richness). Additionally, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples. Importantly, many of these differences are due to undetected chimeras that cannot be fully removed using standard bioinformatic approaches. This is important because it suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.

There were few differences in bacterial community composition based on HiFi DNA polymerase. Additionally, the data suggests that there are no differences in the overall bacterial community composition for sub-sampling depth or cycle number used. One possible reason for this outcome is that our study did not have enough power to detect differences due to low sample number in each group. Another reason is that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence.

Our observations suggest that there are clear HiFi DNA polymerase dependent differences in error rate and chimera prevalence that cannot be fully removed using bioinformatic approaches [Figure 4 & 5]. Although this variation may be due to the DNA polymerase family, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the material safety data sheet (MSDS) the differences between the two HiFi DNA polymerases are not immediately apparent. Both HiFi DNA polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences in chimera rate between the different HiFi DNA polymerases.

@@

The number of OTUs are dependent on both HiFi DNA polymerase and cycle number chosen. Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys because intrinsic differences in HiFi DNA polymerases can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.

### 2018-03-19 submission/manuscript.tex
discussion and conclusion updates

These observations build upon previous studies (Gohl et al., 2016) by
showing that different HiFi DNA polymerases have significant differences
in the number of OTUs and that changes in total OTUs correlate with
chimeras not removed after sequence processing {[}Figure 1-2 \& 5{]}.
HiFi DNA polymerase dependent differences in total number of OTUs are
important to consider since many diversity metrics rely on this measure
as part of their calculations. Our observations show that HiFi DNA
polymerase can have a noticeable affect on the OTUs generated and these
differences are consistent across sub-sampling depth and PCR cycle
number {[}Figure 2-4{]}. Importantly, high biomass samples may
underestimate the biases present within low biomass samples. An
indication that this may be the case is the earlier detection of
differences in total number of OTUs between HiFi DNA polymerases in the
mock versus fecal samples.

Although we did not observe strong differences in the Bray-Curtis index
the data suggests that there may be differences between 15 and 20 cycles
versus higher cycle numbers, such as 30 cycles, that are commonly used.
There was few differences within individuals between corresponding
5-cycle increments (e.g.~15 to 20, 20 to 25, etc.). However, there is a
clear trend that suggests that the community at 20 cycles is different
then the community at 25 cycles {[}Figure 3{]}. These findings, in
conjunction with the PERMANOVA results, suggest that cycle number can
change bacterial community calculations but that these differences are
minimal once 25 cycles are reached. Further, for some HiFI DNA
polymerases, increasing the sub-sampling depth may reduce some of these
observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence
differences between the different HiFi DNA polymerases {[}Figure 5{]}.
The chimera prevalence was strongly correlated with the number of OTUs
which is relied upon heavily for many community metric calculations.
However, Bray-Curtis analysis with PERMANOVA showed few differences
based on HiFi DNA polymerase. It is possible that many of the increased
number of OTUs are not highly abundant allowing the Bray-Curtis index to
be able to successfully down-weight these respective OTUs (Minchin,
1987). The choice of downstream diversity metric could be an important
consideration in helping to mitigate these observed changes due to high
chimera prevalence in specific HiFi DNA polymerases.

Our observations suggest that there are clear HiFi DNA polymerase
dependent differences in both per base error rate and chimera prevalence
that cannot be removed using bioinformatic approaches {[}Figure 4 \&
5{]}. Although it may be a natural assumption that the variation may be
due to the DNA polymerase family, the highest chimera rate, from
Platinum, was a family A polymerase while the lowest, from Accuprime,
was also an A family polymerase (Ishino \& Ishino, 2014). In fact, from
the material safety data sheet (MSDS), the differences between the two
is not immediately apparent. Both Accuprime and Platinum contain a
recombinant \emph{Taq} DNA polymerase, a \emph{Pyrococcus} spp GB-D
polymerase and a platinum \emph{Taq} antibody. It is possible that
differences in how the recombinant \emph{Taq} was generated could be the
main reason for the differences in chimera rate.

>>>>

In this study we show that the number of OTUs, error rate, and chimera
prevalence are HiFi DNA polymerase dependent {[}Figure 1-2 \& 4-5{]}.
These differences are important because many diversity metrics rely on
this measure as part of their calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between HiFi DNA polymerases in the mock versus fecal samples might
indicate that high biomass samples may underestimate the biases present
within low biomass samples. Importantly, many of these differences are
due to undetected chimeras that cannot be fully removed using standard
bioinformatic approaches. This is important because it suggests that
some of the diversity differences between studies can be attributed to
differences in HiFi DNA polyermase used.

There were few differences in bacterial community composition based on
HiFi DNA polymerase. Additionally, the data suggests that there are no
differences in the overall bacterial community composition for
sub-sampling depth or cycle number used. One possible reason for this
outcome is that our study did not have enough power to detect
differences due to low sample number in each group. Another reason is
that many of the OTUs are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight chimeric OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate these observed HiFi DNA
polymerase dependent differences in chimera prevalence.

Our observations suggest that there are clear HiFi DNA polymerase
dependent differences in error rate and chimera prevalence that cannot
be fully removed using bioinformatic approaches {[}Figure 4 \& 5{]}.
Although this variation may be due to the DNA polymerase family, the
highest and lowest chimera rates both belonged to a family A polymerase
(Platinum and Accuprime\textsuperscript{TM} respectively) (Ishino \&
Ishino, 2014). Additionally, based on the material safety data sheet
(MSDS) the differences between the two HiFi DNA polymerases are not
immediately apparent. Both HiFi DNA polymerases contain a recombinant
\emph{Taq} DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a
platinum \emph{Taq} antibody. With everything else being equal, it is
possible that differences in how the recombinant \emph{Taq} was
generated could be a contributing factor for the differences in chimera
rate between the different HiFi DNA polymerases.
